The Spatial and Temporal Domains of Modern Ecology

### 2017-12-19 paper/SI.tex
Basically finished revisions to main manuscript, including responses to reviewers (not versioned). Still need to make one last check through SI.

Of course, we cannot attach confidence to a trend in observational extent extrapolated from a non-significant trend in observational type. However, some additional evidence in support of this possible trend lies within the extent values themselves. The regression analyses applied to dimensions shows an apparent increase in extent by 0.25 orders of magnitude per year (Fig. \ref{sc_by_yr}) between 2004-2014 (R$^2$ = 0.25, p$<$0.07). This somewhat clearer (albeit still not significant at the customary p$<$0.05 threshold) trend suggests that including more recent studies would have increased mean extent, but by a more modest 5.5\% (0.02 orders of magnitude).

>>>>

Of course, we cannot attach confidence to a trend in observational extent extrapolated from a non-significant trend in observational type. However, some additional evidence in support of this possible trend lies within the extent values themselves. The regression analyses applied to dimensions shows an apparent increase in extent of 0.25 orders of magnitude per year (Fig. \ref{sc_by_yr}) between 2004-2014 (R$^2$ = 0.25, p$<$0.07). This weak trend suggests that including more recent studies would have increased mean extent, but by a more modest 5.5\% (0.02 orders of magnitude).

### 2017-12-19 paper/main.tex
Basically finished revisions to main manuscript, including responses to reviewers (not versioned). Still need to make one last check through SI.

\linenumbers
\vspace{10pt}
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992}. This awareness has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad and increasing range of scales \cite{schneider_rise_2001, chave_problem_2013}.

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses of ecological research scales were published nearly 30 years ago \cite{tilman_ecological_1989,kareiva_spatial_1988}, it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}) of empirical observations (defined here as ecological observations collected under un-controlled or non-manipulated conditions) that were reported within recently (2004-2014) published ecological studies. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

\noindent \textbf{\emph{Distributions within the four primary dimensions}}\\
In terms of resolution, the majority (67\%) of observations (across all methods) were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). These distributions primarily reflect those of field observations, the dominant observational methodology. Examining the distributions for each observational method (Fig. S1 in SI) shows that automated sensing and paleo-reconstruction had resolutions that were generally finer (85\% or more $<$0.1 m$^2$) than field observations (47\% $<$0.1 m$^2$), while the majority of remote observations were much coarser (70\% $>$100 m$^2$).

@@

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting 1 year to 1 decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration). The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken; this position

@@

Two primary domains of observational concentration are revealed by juxtaposing duration and extent (across all observations). The first consists of observations spanning one month to one decade in time and 10-1,000 ha in space, while the second is defined by observations of one year to several decades that cover 10,000 to 1,000,000 ha (Fig. 2D).  Three other notable, but lesser areas of concentration are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($<$1 day) of either 1-10 ha or 10,000-1,000,000 ha.  The median observation (1 year duration, 100 ha extent) from automated sensing lies near the center of the second major concentration, while the median extents of remote (1-10 million ha) and field observations bound the primary concentration at its upper and lower extents, while the median duration of both observational types falls between 1 month to 1 year.

@@

Another potential source of bias lies within our scale-estimation protocols, chiefly with respect to our rule for estimating resolution (the smallest areal unit of \emph{complete} measurement). We selected this definition for the sake of consistency, but some papers reported resolution as a larger area in which sub-samples were taken. For these, our estimates are finer than what the studies' authors apparently considered to be plot resolution. Additionally, our domain estimates would presumably be somewhat different if we had included experimentally manipulated observations. For example, average resolution and duration would likely be finer \cite{tilman_ecological_1989,kareiva_spatial_1988}.

@@

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. To evaluate this possibility, we used linear regressions to examine whether the relative frequency of observing methods changed during the 10 year study period, and whether our findings regarding scale domains might be different if we included more recent studies. Although our sample size was too small to assign statistical significance, we found a possible positive trend in the use of remote observations and a corresponding decline in field observations over the 10 year period. If these trends were not spurious, they suggest that including studies from 2015-2017 would result in a somewhat larger relative sample of remote observations, which would in turn increase slightly the mean extent of ecological observations (see SI for details of trend analyses).

@@

Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger plot sizes demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (but shown in Fig. S2 in SI), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these aforementioned tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which line in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

@@

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. A weak positive trend in our own data suggests that average extent of ecological observations is steadily increasing (Fig. S5 in SI), and likely corresponds to the increasing use of remote observation technologies (Fig. S4 in SI).

@@

It is therefore important to examine whether the scales of the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the \emph{scale representativeness} of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. The distance between the adjusted actual value and the ostensible value can provide a measure of how well the replicates represent the intended scale of observation. Though increasing spatial or temporal coverage may not always be the goal of a study (e.g., when spatial or temporal autocorrelation is a measure of interest), if the gap between actual and ostensible values remains large, then alternative sampling methods may be used to close it. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

The aforementioned insights regarding modern observational domains must be tempered by the uncertainty within our own scale estimates, as detailed in the preceding section. However, most of this uncertainty is attributable to unclear reporting of scale values in the majority of papers we reviewed (a problem also noted in geography studies \cite{margulies_ambiguous_2016}). This tendency towards vague documentation offers one final insight, which is that, despite decades of accumulated knowledge regarding the importance of scale in ecology \cite{levin_problem_1992, wiens_spatial_1989, chave_problem_2013, wheatley_factors_2009}, scale appears to remain a low priority throughout much of the discipline. Beyond contributing to the broader problem of scientific reproducibility \cite{baker_1500_2016}, inattentiveness to scale increases the risk that observations inadequately represent the phenomenon of interest, thereby limiting the generalizability of any derived ecological knowledge \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}. To mitigate this problem, we recommend that ecological journals require authors to quantify and clearly report the values of resolution, extent, interval, and duration.

@@

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{../vignettes/figures/fig2.png}
\vspace{-0.15 cm}
\caption{Kernel density estimates of observational densities within the domains defined by A) interval and resolution (of temporally replicated observations only), B) duration and extent, C) resolution and extent, and D) interval and duration (of temporally replicated observations). Density estimates were applied to the log-transformed values of each observational dimension, and density estimates are rescaled to represent percentages. Letters in the plots denote the median values of different observational methods (f=field observations; a = automated sensing; r = remote sensing; p = paleo-observations). The grey shaded areas represent physically impossible domains (intervals greater than duration and resolutions greater than extent).}
\label{afoto1}
\end{figure}

>>>>

\linenumbers
\vspace{10pt}
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992,sandel_scale_2009}. This awareness has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad and increasing range of scales \cite{schneider_rise_2001, chave_problem_2013}.

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses \cite{tilman_ecological_1989,kareiva_spatial_1988} of ecological research scales were published nearly 30 years ago (but see \cite{porter_crop_2005,sandel_scale_2009} for more recent reviews), it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}) of empirical observations (defined here as ecological observations collected under un-controlled or non-manipulated conditions) that were reported within recently (2004-2014) published ecological studies. Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

\noindent \textbf{\emph{Distributions within the four primary dimensions}}\\
In terms of resolution, spatial replicates for the majority (67\%) of observations (across all methods) were $<$1 m$^2$, a further 24\% were 1 m$^2$ up to 1 ha, and the remaining 9\% were $\geq$1 ha (Fig. 1A). These distributions primarily reflect those of field observations, the dominant observational methodology. Examining the distributions for each observational method (Fig. S1 in SI) shows that automated sensing and paleo-reconstruction had resolutions that were generally finer (85\% or more $<$0.1 m$^2$) than field observations (47\% $<$0.1 m$^2$), while the majority of remote observations were much coarser (70\% $>$100 m$^2$).

@@

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting 1 year to 1 decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration).

@@

Two primary domains of observational concentration are revealed by juxtaposing duration and extent (across all observations). The first consists of observations spanning one month to one decade in time and 10-1,000 ha in space, while the second is defined by observations of one year to several decades that cover 10,000 to 1,000,000 ha (Fig. 2D).  Three other notable, but lesser areas of concentration are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($<$1 day) of either 1-10 ha or 10,000-1,000,000 ha.  The median observation (1 year duration, 100 ha extent) from automated sensing lies near the center of the second major concentration, while the median extents of field (1-10,000 ha) and remote (1-10 million ha) observations bound the primary concentration at its upper and lower extents, while the median duration of both observational types falls between 1 month to 1 year.

@@

Another potential source of bias lies within our scale-estimation protocols, chiefly with respect to our rule for estimating resolution (the smallest areal unit of \emph{complete} measurement). We selected this definition for the sake of consistency, but some papers reported resolution as a larger area in which sub-samples were taken. For these, our estimates are finer than what the studies' authors apparently considered to be the resolution of spatial replicates. Our domain estimates would presumably be somewhat different if we had included experimentally manipulated observations. For example, average resolution and duration would likely be finer \cite{tilman_ecological_1989,kareiva_spatial_1988}. Additionally, the token 1 second (SI) we used to represent the duration of remotely sensed temporal replicates (which are effectively instantaneous) mean that we underestimated actual duration-duration differences for this observational method (Fig. S3). However, the relatively low proportion of remote observations suggests that the impact on our overall findings was negligible (Fig. 3).

@@

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. To evaluate this possibility, we used linear regressions to examine whether the relative frequency of observing methods changed during the 10 year study period, and whether our findings regarding scale domains might be different if we included more recent studies. Although our sample size was too small to assign statistical significance, we found a possible positive trend in the use of remote observations and a corresponding decline in field observations over the study period. If these trends were not spurious, they suggest that including studies from 2015-2017 would result in a somewhat larger relative sample of remote observations, which would in turn slightly increase the mean extent of ecological observations (see SI for details of trend analyses).

@@

Contrasting observational dimensions reveals that larger extents are associated with larger spatial replicates (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger spatial replicates demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (but shown in Fig. S2 in SI), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these aforementioned tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which lie in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

@@

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. A weak positive trend in our own data also suggests that average extent of ecological observations is steadily increasing (Fig. S5 in SI), and likely corresponds to the increasing use of remote observation technologies (Fig. S4 in SI).

@@

It is therefore important to examine whether the scales of the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the \emph{scale representativeness} of replicates, which is to i) calculate the autocorrelation (spatial or temporal) within the observations (using a semi-variogram, for example), ii) find the threshold distance (or time) below which a suitably strong correlation (e.g. r=0.7) will exist between neighboring sampled values, iii) add that distance (or time) to the sample resolution (or duration), and iv) recalculate actual extent (or duration) using the adjusted resolution (or sampling duration). The difference between this autocorrelation-adjusted actual extent (or actual duration) and extent (or duration) may provide a useful additional measure of how well the replicates represent the intended scale of observation. Though increasing spatial or temporal coverage may not always be the goal of a study (e.g., when spatial or temporal autocorrelation is a measure of interest), if the gap between actual and ostensible values remains large, then alternative sampling methods may be used to close it. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

The aforementioned insights regarding modern observational domains must be tempered by the uncertainty within our own scale estimates, as detailed in the preceding section. However, most of this uncertainty is attributable to unclear reporting of scale values in the majority of papers we reviewed (a problem also noted in geography studies \cite{margulies_ambiguous_2016}). This tendency towards vague documentation offers one final insight, which is that, despite decades of accumulated knowledge regarding the importance of scale in ecology \cite{levin_problem_1992, wiens_spatial_1989, chave_problem_2013, wheatley_factors_2009}, scale appears to remain a low priority throughout much of the discipline. Beyond contributing to the broader problem of scientific reproducibility \cite{baker_1500_2016}, inattentiveness to scale increases the risk that observations inadequately represent the phenomenon of interest, thereby limiting the generalizability of any derived ecological knowledge \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}. To mitigate this problem, we recommend that ecological journals require authors to quantify and clearly report the values of resolution, extent, interval, and duration. Fortunately, some journals already appear to be implementing such policies. For example, \emph{Global Ecology and Biogeography} now requires information on the spatial, temporal and taxonomic scale of studies to be in the abstract (a policy adopted in early 2016).

@@

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{../vignettes/figures/fig2.png}
\vspace{-0.15 cm}
\caption{Kernel density estimates of observational densities within the domains defined by A) interval and resolution (of temporally replicated observations only), B) duration and extent, C) resolution and extent, and D) interval and duration (of temporally replicated observations). Density estimates were applied to the log-transformed values of each observational dimension, and density estimates are rescaled to represent percentages. Letters in the plots denote the median values of different observational methods (f=field observations; a = automated sensing; r = remote sensing; p = paleo-observations). The grey shaded areas represent physically impossible domains (intervals greater than duration and resolutions greater than extent). Density values below the lower 3rd percentile fall within the darkest portion of the color gradient.}
\label{afoto1}
\end{figure}
